---
title: "Part2_Outlier_Detection"
author: "Coline Caillon"
date: "2026-09-15"
output: github_document
---

# Detect candidate loci under selection

In 'RISG_dDocent_working' directory, make new directory for outlier detection and link vcf files and popmap file.
```{bash}
cd /home/ccaillon/RISG_dDocent_working
mkdir RISG_Outlier_Detection_working
cd RISG_Outlier_Detection_working

cp ../RISG_SNP_Filtering_working/SNP.TRSdp5g75mafPM9g9maf1.2A.nDNA.FIL.recode.vcf .
cp ../RISG_SNP_Filtering_working/popmap .
```

## 1. pcadapt

Assumes that genetic variants (SNPs) that are excessively correlated with population structure are outliers, making them strong candidates for being under natural selection. It first captures the overarching population structure by performing PCA on centered and scaled genotype data. It infers structure directly from the genetic data, doesn't required to group individuals into explicit populations beforehand.

Load in R:
```{r}
# Install and load pcadapt
#install.packages("pcadapt")
library(pcadapt)
```

```{r}
# Convert VCF file to pcadapt format
vcf2pcadapt("SNP.TRSdp5g75mafPM9g9maf1.2A.nDNA.FIL.recode.vcf", output = "SNP.pcadapt", allele.sep = c("/", "|"))
filename <- read.pcadapt("SNP.pcadapt", type = "pcadapt")
```

```{r}
# Choosing the number K of Principal Components
# We start off with a large number of PCs
RISGpops <- pcadapt(input = filename, K = 20)
```

```{r}
# Plot the likelihoods using a scree plot
plot(RISGpops, option = "screeplot")
```

```{r}
# Plotting likelihoods again, but with 10 Principal Components
plot(RISGpops, option = "screeplot", K = 10)
```

```{r}
# Create population designations
poplist.names <- c(rep("BAR", 10), rep("BIS", 10), rep("CS", 20), rep("GB", 10), rep("HA", 20),
                   rep("HC", 20), rep("HI", 20), rep("HO", 20), rep("KIC", 10), rep("MCD", 9),
                   rep("M", 20), rep("MW", 20), rep("NB", 20), rep("PVD", 10), rep("SR", 20))
```

```{r}
# Create estuary designations
estuarylist.names <- ifelse(poplist.names %in% c("BAR", "BIS", "GB", "KIC", "MCD", "PVD"), "NB",
                            ifelse(poplist.names %in% c("CS", "HC", "M", "NB", "SR"), "DB",
                                   ifelse(poplist.names %in% c("HA", "HI", "HO", "MW"), "CB",
                                          NA)))
```

```{r}
library(ggplot2)

pca.pop <- as.data.frame(RISGpops$score)
pca.pop$Populations <- poplist.names

# Define colors and shapes for each population
pop_fills <- c("BIS"="#004700", "MCD"="#006b00", "GB"="#008f00", "KIC"="#19ba19", "BAR"="#7fd97f","PVD"="#ccefcc",
                "CS"="#00007f", "NB"="#0000bf", "SR"="#0000ff", "M"="#5959ff", "HC"="#b2b2ff",
                "HI"="#990000", "HO"="#d90000", "HA"="#ff3232", "MW"="#ff7373")
pop_colors <- c("BIS"="white", "MCD"="white", "GB"="white", "KIC"="white", "BAR"="white","PVD"="white",
                "CS"="white", "NB"="white", "SR"="white", "M"="white", "HC"="white",
                "HI"="white", "HO"="white", "HA"="white", "MW"="white")
pop_shapes <- c("BIS"=21, "MCD"=21, "GB"=21, "KIC"=21, "BAR"=21,"PVD"=21,
                "CS"=21, "NB"=21, "SR"=21, "M"=21, "HC"=21,
                "HI"=21, "HO"=21, "HA"=21, "MW"=21)

# Make score plot
ggplot(pca.pop, aes(x = V1, y = V2, color = Populations, fill = Populations, shape = Populations)) +
  geom_point(size = 3) +
  scale_fill_manual(values = pop_fills) +
  scale_color_manual(values = pop_colors) +
  scale_shape_manual(values = pop_shapes) +
  theme(panel.background = element_blank(),
        panel.border = element_rect(color="black", fill=NA, linewidth=0.5),
        panel.grid.major = element_line(color="gray90", linewidth=0.15),
        panel.grid.minor = element_blank(),
        axis.line = element_line(0.2, colour = "black"),
        axis.text.x = element_text(color="black", size=9),
        axis.text.y = element_text(color="black", size=9),
        axis.ticks = element_line(color = "black"),
        axis.title.x = element_text(color="black", size=9),
        axis.title.y = element_text(color="black", size=9),
        legend.position = "none") +
  labs(x = "PC1", y = "PC2")
```

```{r}
# Score Plot
# Plot the first two PCs
plot(RISGpops, option = "scores", pop = poplist.names)
```

```{r}
# PCA plot with PC3 and 4
plot(RISGpops, option = "scores", i = 3, j = 4, pop = estuarylist.names)
```

Evaluating LD in dataset -> looking to see if loadings are clustered in certain genomic regions (following https://bcm-uga.github.io/pcadapt/articles/pcadapt.html#f--linkage-disequilibrium-ld-thinning)

Linkage disequilibrium (LD) = non-random association of alleles at two or more different loci in a population, meaning certain combinations of alleles are inherited together more or less often than expected by chance.

Inversion -> unusually strong/extended LD across a large genomic region.
Clues of inversion: PCA shows 3 discrete clusters and the SNPs driving the separation are concentrated on one chromosome.

```{r}
par(mfrow = c(2, 2))
for (i in 1:4)
  plot(RISGpops$loadings[, i], pch = 19, cex = .3, ylab = paste0("Loadings PC", i))
```

