---
title: "Part2_Outlier_Detection"
author: "Coline Caillon"
date: "2026-09-15"
output: github_document
---

# Detect candidate loci under selection

In 'RISG_dDocent_working' directory, make new directory for outlier detection and link vcf files and popmap file.
```{bash}
cd /home/ccaillon/RISG_dDocent_working
mkdir RISG_Outlier_Detection_working
cd RISG_Outlier_Detection_working

cp ../RISG_SNP_Filtering_working/SNP.TRSdp5g75mafPM9g9maf1.2A.nDNA.FIL.recode.vcf .
cp ../RISG_SNP_Filtering_working/popmap .
```

## 1. pcadapt

Assumes that genetic variants (SNPs) that are excessively correlated with population structure are outliers, making them strong candidates for being under natural selection. It first captures the overarching population structure by performing PCA on centered and scaled genotype data. It infers structure directly from the genetic data, doesn't required to group individuals into explicit populations beforehand.

Load in R:
```{r}
# Install and load pcadapt
#install.packages("pcadapt")
library(pcadapt)
```

```{r}
# Convert VCF file to pcadapt format
vcf2pcadapt("SNP.TRSdp5g75mafPM9g9maf1.2A.nDNA.FIL.recode.vcf", output = "SNP.pcadapt", allele.sep = c("/", "|"))
filename <- read.pcadapt("SNP.pcadapt", type = "pcadapt")
```



