---
title: "Part2_Outlier_Detection"
author: "Coline Caillon"
date: "2026-09-15"
output: github_document
---

# Detect candidate loci under selection

In 'RISG_dDocent_working' directory, make new directory for outlier detection and link vcf files and popmap file.
```{bash}
cd /home/ccaillon/RISG_dDocent_working
mkdir RISG_Outlier_Detection_working
cd RISG_Outlier_Detection_working

cp ../RISG_SNP_Filtering_working/SNP.TRSdp5g75mafPM9g9maf1.2A.nDNA.FIL.recode.vcf .
cp ../RISG_SNP_Filtering_working/popmap .
```

## 1. pcadapt

Assumes that genetic variants (SNPs) that are excessively correlated with population structure are outliers, making them strong candidates for being under natural selection. It first captures the overarching population structure by performing PCA on centered and scaled genotype data. It infers structure directly from the genetic data, doesn't required to group individuals into explicit populations beforehand.

Load in R:
```{r}
# Install and load pcadapt
#install.packages("pcadapt")
library(pcadapt)
```

```{r}
# Convert VCF file to pcadapt format
vcf2pcadapt("SNP.TRSdp5g75mafPM9g9maf1.2A.nDNA.FIL.recode.vcf", output = "SNP.pcadapt", allele.sep = c("/", "|"))
filename <- read.pcadapt("SNP.pcadapt", type = "pcadapt")
```

```{r}
# Choosing the number K of Principal Components
# We start off with a large number of PCs
RISGpops <- pcadapt(input = filename, K = 20)
```

```{r}
# Plot the likelihoods using a scree plot
plot(RISGpops, option = "screeplot")
```
Cattell’s rule indicates K = 6.

```{r}
# Create population designations
poplist.names <- c(rep("BAR", 10), rep("BIS", 10), rep("CS", 20), rep("GB", 10), rep("HA", 20),
                   rep("HC", 20), rep("HI", 20), rep("HO", 20), rep("KIC", 10), rep("MCD", 9),
                   rep("M", 20), rep("MW", 20), rep("NB", 20), rep("PVD", 10), rep("SR", 20))
```

```{r}
# Create estuary designations
estuarylist.names <- ifelse(poplist.names %in% c("BAR", "BIS", "GB", "KIC", "MCD", "PVD"), "NB",
                            ifelse(poplist.names %in% c("CS", "HC", "M", "NB", "SR"), "DB",
                                   ifelse(poplist.names %in% c("HA", "HI", "HO", "MW"), "CB",
                                          NA)))
```

```{r}
library(ggplot2)

pca.pop <- as.data.frame(RISGpops$score)
pca.pop$Populations <- poplist.names

# Define colors and shapes for each population
pop_fills <- c("BIS"="#004700", "MCD"="#006b00", "GB"="#008f00", "KIC"="#19ba19", "BAR"="#7fd97f","PVD"="#ccefcc",
                "CS"="#00007f", "NB"="#0000bf", "SR"="#0000ff", "M"="#5959ff", "HC"="#b2b2ff",
                "HI"="#990000", "HO"="#d90000", "HA"="#ff3232", "MW"="#ff7373")
pop_colors <- c("BIS"="white", "MCD"="white", "GB"="white", "KIC"="white", "BAR"="white","PVD"="white",
                "CS"="white", "NB"="white", "SR"="white", "M"="white", "HC"="white",
                "HI"="white", "HO"="white", "HA"="white", "MW"="white")
pop_shapes <- c("BIS"=21, "MCD"=21, "GB"=21, "KIC"=21, "BAR"=21,"PVD"=21,
                "CS"=21, "NB"=21, "SR"=21, "M"=21, "HC"=21,
                "HI"=21, "HO"=21, "HA"=21, "MW"=21)

# Make score plot
ggplot(pca.pop, aes(x = V1, y = V2, color = Populations, fill = Populations, shape = Populations)) +
  geom_point(size = 3) +
  scale_fill_manual(values = pop_fills) +
  scale_color_manual(values = pop_colors) +
  scale_shape_manual(values = pop_shapes) +
  theme(panel.background = element_blank(),
        panel.border = element_rect(color="black", fill=NA, linewidth=0.5),
        panel.grid.major = element_line(color="gray90", linewidth=0.15),
        panel.grid.minor = element_blank(),
        axis.line = element_line(0.2, colour = "black"),
        axis.text.x = element_text(color="black", size=9),
        axis.text.y = element_text(color="black", size=9),
        axis.ticks = element_line(color = "black"),
        axis.title.x = element_text(color="black", size=9),
        axis.title.y = element_text(color="black", size=9),
        legend.position = "none") +
  labs(x = "PC1", y = "PC2")
```

```{r}
# Score Plot
# Plot the first two PCs
plot(RISGpops, option = "scores", pop = poplist.names)
```

```{r}
# PCA plot with PC3 and 4
plot(RISGpops, option = "scores", i = 3, j = 4, pop = estuarylist.names)
```

Evaluating LD in dataset -> looking to see if loadings are clustered in certain genomic regions (following https://bcm-uga.github.io/pcadapt/articles/pcadapt.html#f--linkage-disequilibrium-ld-thinning)

Linkage disequilibrium (LD) = non-random association of alleles at two or more different loci in a population, meaning certain combinations of alleles are inherited together more or less often than expected by chance.
-> can affect ascertainment of population structure, dense SNP data should account for LD in their genome scans.

Inversion -> unusually strong/extended LD across a large genomic region = regions of low recombination.
Clues of inversion: PCA shows 3 discrete clusters and the SNPs driving the separation are concentrated on one chromosome.

```{r}
# To evaluate if LD might be an issue for dataset, display the loadings (contributions of each SNP to the PC) and evaluate if the loadings are clustered in a single or several genomic regions.
par(mfrow = c(2, 3))
for (i in 1:6)
  plot(RISGpops$loadings[, i], pch = 19, cex = .3, ylab = paste0("Loadings PC", i))
```
We see that PC2 is determined by two genomic regions of strong LD, also maybe one region in PC1.
We should therefore thin SNPs in order to compute the PCS = apply LD thinning.
I'm going to use the default values for window size and r2 threshold for pcadapt (size = 200, thr = 0.1 -> good general starting point and conservative choice for thr).
However I still see these two strong LD regions in PC2, and now also in PC4 -> I will modify parameters of SNP thinning.
- size = window size -> the number of SNPs considered around each SNP, not a physical distance in bp (size = 500 more aggressive genomic thinning if strong/long-range LD).
- thr = r² threshold -> SNPs with LD above this threshold are thinned so that highly correlated markers don't disproportionately influence PCA (thr = 0.05 very stringent if strong LD or very dense data).

```{r}
# Performing SNP thinning in order to compute PCs
res <- pcadapt(filename, K = 20, LD.clumping = list(size = 1000, thr = 0.1))
plot(res, option = "screeplot")
```

After SNP thinning, we choose K = 4.

```{r}
res <- pcadapt(filename, K = 4, LD.clumping = list(size = 500, thr = 0.1))
par(mfrow = c(2, 2))
for (i in 1:4)
  plot(res$loadings[, i], pch = 19, cex = .3, ylab = paste0("Loadings PC", i))
```

```{r}
plot(res, option = "manhattan")
```

```{r}
res$singular.values
```

Choosing a cutoff for outlier detection
```{r}
library(qvalue) #transforms p-values into q-values

qval <- qvalue(res$pvalues)$qvalues
alpha <- 0.1 #expected false discovery rate lower than 10%

# Save outliers
pcadapt.outliers <- which(qval < alpha)
length(pcadapt.outliers)

```

7141 outliers
